---
title: "Example code for dimensionality reduction and clustering in R"
author: "Kevin Rue-Albrecht"
date: "03/10/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(cowplot)
library(umap)
library(Rtsne)
library(dendextend)
library(dbscan)
```

# Exercise

## Setup

- Import the `iris` data set.

```{r}
iris <- iris
head(iris)
```

- Separate the matrix of measurements in a new object named `iris_features`.

```{r}
iris_features <- iris[,c(1:4)]
head(iris_features)

iris_features <- as.matrix(iris_features)

```

# Exercise

## Apply Principal Components Analysis (PCA)

The `prcomp()` function allows you to standardise the data as part of the principal components analysis itself.

- Apply PCA while centering and scaling the matrix of features.

```{r}
pca_iris <- prcomp(iris_features, center = T, scale. = T)
pca_iris
```

- Examine the PCA output.
  Display the loading of each feature on each principal component.
  The rotation value gives the weigh or loading for each PC

```{r}
str(pca_iris)
pca_iris$rotation 
```

```{r}

```

- Use the return value of the PCA to create a `data.frame` called `pca_iris_dataframe` that contains the coordinates projected on principal components.

```{r}
#we know that x is the sample position after rotation

pca_iris_dataframe <- as.data.frame(pca_iris$x)
pca_iris_dataframe <- cbind(pca_iris_dataframe, 'Species'=iris$Species)
head(pca_iris_dataframe)

```

- Visualise the PCA projection using `ggplot2::geom_point()`.

```{r}
ggplot(pca_iris_dataframe, aes(x=PC1, y=PC2, color=Species)) +
  geom_point()
  
  
```

### Bonus point

- Color data points according to their class label.

- Store the PCA plot as an object named `pca_iris_species`.

```{r}
 pca_iris_species <- ggplot(pca_iris_dataframe, aes(x=PC1, y=PC2, color=Species)) +
  geom_point()

```

```{r}
pca_iris_species <- ggplot(pca_iris_dataframe,    ) +
  
  
pca_iris_species
```

# Exercise

## Variable loading

- Color a scatter plot of PC1 and PC2 by the value of the variable most strongly associated with the first principal component.

Find this from rotation matrix. 
Then add a column of the variable giving rise to PC1, plot this as the colour of the ggplot.

What do you observe?

```{r}
pca_iris$rotation

pca_iris_dataframe_drivingvariable <- cbind(pca_iris_dataframe, Petal.Length=iris$Petal.Length)

head(pca_iris_dataframe_drivingvariable)
```

```{r}
 ggplot(pca_iris_dataframe_drivingvariable, aes(x=PC1, y=PC2, color=Petal.Length, shape=Species)) +
  geom_point(size=3)

```

> Answer:
> 
> 

## Variance explained

- Compute the variance explained by principal components, using information present in the return value of the `prcomp()` function.

```{r}
sdev <- pca_iris$sdev


sdsq <- pca_iris$sdev^2
sumsdsq <- sum(sdsq)

sdev <- pca_iris$sdev

explained_variance_ratio <- sdev^2/sumsdsq
explained_variance_ratio
```

- Visualise the variance explained by each principal component using `ggplot2::geom_col()`.

```{r}
names(explained_variance_ratio) <- c('PC1','PC2','PC3','PC4')

explained_variance_ratio

variance_dataframe <- as.data.frame(explained_variance_ratio)

head(variance_dataframe)
#this has only one column and rownames - need to change rownames to a column.

variance_dataframe$name <- rownames(variance_dataframe)


#alternative:  
#variance_dataframe <- data.frame(name=names(explained_variance_ratio), variane=explained_variance_ratio)

#names(explained_variance_ratio) <- paste0('pPC', 1:length(explained_variance_ratio))
```

```{r}
ggplot(variance_dataframe,  aes(x=name, y=explained_variance_ratio)) +
  geom_col()+
  labs(x='PCA1=72.9%', y='PCA2=22.8%')

  
  
  
```

# Exercise

## UMAP

- Apply UMAP on the output of the PCA.

```{r}
set.seed(1) # Set a seed for reproducible results

#use pca to reduce dimensions before putting into umap.  Doing umap directly on gene expression dataset not wise due to data intensity of computing
umap_iris <- umap(pca_iris$x)
#don't use this as give v big table head(umap_iris)
```

- Inspect the UMAP output.

```{r}
str(umap_iris)
head(umap_iris$layout)
names(umap_iris$layout) <- c('X1','X2')

```

- Visualise the UMAP projection using `ggplot2::geom_point()`.

```{r}
umap_iris_dataframe <- as.data.frame(umap_iris$layout)
head(umap_iris_dataframe)

umap_iris_dataframe <- cbind(umap_iris_dataframe, Species=iris$Species)

ggplot(umap_iris_dataframe, aes(x=V1, y=V2, colour=Species))+
  geom_point()

head(umap_iris_dataframe)
```

```{r}
ggplot(umap_iris_dataframe, aes(x=)) +

  
```

### Bonus point

- Color data points according to their class label.

- Store the UMAP plot as an object named `umap_iris_species`.

```{r}
umap_iris_species <- ggplot(umap_iris_dataframe, aes(x=V1, y=V2, colour=Species))+
  geom_point()

umap_iris_species

head(umap_iris_dataframe)
```

# Exercise

## t-SNE

- Apply t-SNE and inspect the output.

```{r}
set.seed(1) # Set a seed for reproducible results
tsne_iris <- Rtsne(pca_iris$x, check_duplicates = FALSE, pca = FALSE, pca_center = F)
str(tsne_iris)
```

- Use the return value of the t-SNE to create a `data.frame` called `tsne_iris_dataframe` that contains the coordinates.

```{r}
head(tsne_iris$Y)
dim(tsne_iris$Y)
class(tsne_iris$Y)
tsne_iris_dataframe <- as.data.frame(tsne_iris$Y)
dim(tsne_iris_dataframe)
head(tsne_iris_dataframe)
```

- Visualise the t-SNE projection.

```{r}
ggplot(tsne_iris_dataframe, aes(x=V1, y=V2)) +
  geom_point()
  
  
```

### Bonus points

- Color data points according to their class label.

- Store the t-SNE plot as an object named `tsne_iris_species`.

```{r}
tsne_iris_dataframe <- cbind(tsne_iris_dataframe,Species=iris$Species)
ggplot(tsne_iris_dataframe, aes(x=V1, y=V2, colour=Species)) +
  geom_point()

head(tsne_iris_dataframe)
```

```{r}
tsne_iris_species <- ggplot(tsne_iris_dataframe, aes(x=V1, y=V2, colour=Species)) +
  geom_point()
  
  
tsne_iris_species
```

- Combine PCA, UMAP and t-SNE plots in a single figure.

```{r, fig.height=6, fig.width=6}
cowplot::plot_grid(pca_iris_species,umap_iris_species, tsne_iris_species, labels=c('PCA', 'UMAP','tSNE'), ncol=2
                   )
  
  
  
  
)
```

# Exercise

## Hierarchical clustering

- Perform hierarchical clustering on the `iris_features` data set,
  using the `euclidean` distance and method `ward.D2`.
  Use the functions `dist()` and `hclust()`.

```{r}
dist_iris <- dist(iris_features, method='euclidean')
hclust_iris_ward <- hclust(dist_iris, method = 'ward.D2')
hclust_iris_ward
```

- Plot the clustering tree.
  Use the function `plot()`.

```{r}
plot(hclust_iris_ward)
```

How many clusters would you call from a visual inspection of the tree?

> Answer:3
> 
> 

- **Bonus point:** Color leaves by known species (use `dendextend`).

```{r}
iris_hclust_dend <- as.dendrogram(hclust_iris_ward)
labels_colors(iris_hclust_dend) <- as.numeric(iris$Species)
plot(iris_hclust_dend)
```

- Cut the tree in 3 clusters and extract the cluster label for each flower.
  Use the function `cutree()`.

```{r}
iris_hclust_ward_labels <- cutree(hclust_iris_ward, k=3 )
iris_hclust_ward_labels
clustered_iris <- cbind(iris, Cluster=iris_hclust_ward_labels)
```


- Repeat clustering using 3 other agglomeration methods:

  + `complete`
  + `average`
  + `single`

```{r}
# complete
hclust_iris_complete <- hclust(dist_iris, method = 'complete')
iris_hclust_complete_labels <- cutree(hclust_iris_complete, k=3)
iris_hclust_complete_labels
```

```{r}
# average
hclust_iris_average <- hclust(dist_iris, method = 'average')
iris_hclust_average_labels <- cutree(hclust_iris_average, k=3)
iris_hclust_average_labels
```

```{r}
# single
hclust_iris_single <- hclust(dist_iris, method = 'single')
iris_hclust_single_labels <- cutree(hclust_iris_single, k=3)
iris_hclust_single_labels
```

- Compare clustering results on scatter plots of the data.

```{r}
iris_clusters_dataframe <- iris
iris_clusters_dataframe$hclust_average <- as.factor(iris_hclust_average_labels)
iris_clusters_dataframe$hclust_complete <- as.factor(iris_hclust_complete_labels)
iris_clusters_dataframe$hclust_single <- as.factor(iris_hclust_single_labels)
iris_clusters_dataframe$hclust_ward <- as.factor(iris_hclust_ward_labels)
iris_clusters_dataframe
```

```{r, fig.height=8, fig.width=10}
plot_average <- ggplot(iris_clusters_dataframe, aes(x=Sepal.Length, y=Sepal.Width, color=hclust_average)) +
  geom_point()
plot_average
  
  
plot_complete <- ggplot(iris_clusters_dataframe, aes(x=Sepal.Length, y=Sepal.Width, color=hclust_complete)) +
  geom_point()
plot_complete

plot_single <- ggplot(iris_clusters_dataframe,aes(x=Sepal.Length, y=Sepal.Width, color=hclust_single)) +
  geom_point()
plot_single
  
  
plot_ward <- ggplot(iris_clusters_dataframe, aes(x=Sepal.Length, y=Sepal.Width, color=hclust_ward))+
  geom_point()
plot_ward
  
library(cowplot)
cowplot_hclus<- cowplot::plot_grid(plot_average,plot_complete,plot_single,plot_ward, ncol=2,nrow=2)

ggsave('Hierarchical_clustering.pdf', cowplot_hclus,height=8, width=10, units='cm')
  

```

# Exercise

## dbscan

- Apply `dbscan` to the `iris_features` data set.

```{r}
dbscan_iris <- dbscan(iris_features, eps=0.5, minP=5)
dbscan_iris
```

- Visualise the `dbscan` cluster label on a scatter plot of the data.

```{r}
iris_clusters_dataframe$dbscan <- as.factor(dbscan_iris$cluster)
head(iris_clusters_dataframe)

```

```{r}


 ggplot(iris_clusters_dataframe, aes(x=Sepal.Length, y=Sepal.Width, color=dbscan, shape=hclust_ward)) +
  geom_point()
  
  
dbscan_plot <- ggplot(iris_clusters_dataframe, aes(x=Sepal.Length, y=Sepal.Width, color=dbscan, shape=hclust_ward)) +geom_point()

dbscan_plot

#dbscan group 0 are the outliers.
```

## hdbscan

- Apply `hdbscan` to the `iris_features` data set.

```{r}
hdbscan_iris <- hdbscan(iris_features, minPts = 5)
hdbscan_iris
```

- Visualise the `hdbscan` cluster label on a scatter plot of the data.

```{r}
iris_clusters_dataframe$hdbscan <- as.factor(hdbscan_iris$cluster)
head(iris_clusters_dataframe)
```

```{r}
hdbscan_plot <- ggplot(iris_clusters_dataframe, aes(x=Sepal.Length, y=Sepal.Width, colour=hdbscan)) +
  geom_point()
  
  
hdbscan_plot
```

## Bonus point

- Combine the plots of `dbscan` and `hdbscan` into a single plot.

```{r, fig.height=3, fig.width=6}
cowplot::plot_grid(
  
  
  
)
```

# Exercise

## K-means clustering

- Apply $K$-means clustering with $K$ set to 3 clusters.

```{r}
set.seed(1) # Set a seed for reproducible results
kmeans_iris <- kmeans(   )
kmeans_iris
```

- Inspect the output.

```{r}

```

- Extract the cluster labels.

```{r}

```

- Extract the coordinates of the cluster centers.

```{r}

```

- Construct a data frame that combines the `iris` dataset and the cluster label.

```{r}
iris_labelled <- iris
iris_labelled$Kmeans <- as.factor(   )
head(iris_labelled)
```

- Plot the data set as a scatter plot.

  + Color by cluster label.

```{r}
ggplot(iris_labelled,    ) +
  
  
```

### Bonus point

- Add cluster centers as points in the plot.

```{r}
iris_means_centers <- as.data.frame(   )
iris_means_centers$Kmeans <- as.factor(   )
head(iris_means_centers)
```


```{r}
ggplot(iris_labelled,    ) +
  
  
  
```

# Exercise

## Cross-tabulation with ground truth

- Cross-tabulate cluster labels with known labels.

```{r}
table(   )
```

How many observations are mis-classified by $K$-means clustering?

> Answer:
> 
> 
> 
> 
> 

## Elbow plot

- Plot the "total within-cluster sum of squares" for K ranging from 2 to 10.

```{r}

```

```{r}
get_mean_totss_for_k <- function(k, data) {
  kmeans_out <- kmeans(data, k)
  return(kmeans_out$tot.withinss)
}
k_range <- 2:10
kmean_totwithinss <- vapply(   )
kmean_totwithinss
```

```{r}
kmean_totwithinss_dataframe <- data.frame(
  K = ,
  totss = 
)
head(kmean_totwithinss_dataframe)
```

```{r}
ggplot(kmean_totwithinss_dataframe,    ) +
  
  
  
```

Do you agree that 3 is the optimal number of clusters for this data set?

> Answer:
> 
> 
> 
> 

